---
title: "Predicting Grasshopper Sparrow Abundance by Landscape Characteristics"
author: "Daniel Redwine"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true # table of content true
    toc_depth: 1  # up to one depth of headings, no subheader (specified by # in html)
    number_sections: false  ## if you want number sections at each table header
    theme: united  # many options for theme
---
```{r setup, echo=FALSE, warning=FALSE, include = FALSE}
knitr::opts_chunk$set(fig.asp = .66, fig.width = 5)
#Its common to have a setup area to help keep track of required packages and formatting options

library(tidyverse)  #importing, tidying, plotting data
library(knitr)      #making tables
library(leaflet)
library(dplyr)
library(hexbin)
library(ggplot2)
library(readr)
library(RColorBrewer)
library(ggfortify)
library(tidyr)
library(gridExtra)
library(gtsummary)
library(gt)
library(flextable)
library(arm)
library(performance)
library(AICcmodavg)
library(grid)

#knitr::opts_chunk$set(fig.width = 5, fig.asp = 1/3) #force figures to be a certain size and aspect ratio
#opts_chunk$set(echo=FALSE)   #can force hide all code in the output; disables 'code folding'
```
# Grasshopper Sparrow Abundance

Grasshopper Sparrows are a grassland obligate species with a distribution across much of the continental United States. Using a dataset obtained from Dryad [here](https://doi.org/10.5061/dryad.9pr1476), we will model Grasshopper Sparrow abundance based on landscape characteristics. 

<center>
<iframe src="https://macaulaylibrary.org/asset/335564991/embed" height="533" width="640" frameborder="0" allowfullscreen></iframe>
</center>
<center> Figure 1. Photograph of a Grasshopper Sparrow </center>

```{r load dataset, message=FALSE, echo=FALSE, error=FALSE}

grsp_abundance_data <- read.csv("../data/grsp_glmm_input_data.csv")

grsp_abundance_data_clean <- grsp_abundance_data %>%
  dplyr::filter(visit == 2) %>% # filtering out visit to avoid random effect
  dplyr::filter(year == 2015) %>% # filtering out year to avoid random effect
  dplyr::select(grsp, ca_400, grass_tree_ed_400)

grsp_abundance_binary_data <- grsp_abundance_data_clean %>%
  mutate(grsp = if_else(grsp >= 1, 1, 0))

grsp_abundance_nonzero_data <- grsp_abundance_data_clean %>%
  filter(grsp >= 1)
  

```

# Visualizing Abundance Data

```{r abundance histogram, warning=FALSE, echo=FALSE, message=FALSE, fig.show = "hold", out.width = "85%", fig.align = "center", fig.cap= "Figure 2. Histogram showing counts of different grasshopper sparrow abundances"}
ggplot(grsp_abundance_data_clean, aes(x=grsp)) + geom_histogram(binwidth=1, colour="black", fill="skyblue")+
  xlab("Grasshopper Sparrow Abundance") +
  ylab("Count")+
  theme_bw()

```


# Poisson GLM

Total grassland area within 400 meters

# Separating Zeroes

```{r binomial model plot, message=FALSE, echo=FALSE, warning=FALSE, fig.show = "hold", out.width = "85%", fig.align = "center", fig.cap= "Figure 3. Model of probability of occurrence for a given core area of grassland (ha) within 400 m"}
grsp_ha_binomial_graph <- ggplot(grsp_abundance_binary_data, aes(ca_400, grsp)) +
  geom_count(show.legend = FALSE) + # Geom count changes size of points to count
  geom_smooth(method="glm", method.args=list(family="binomial"(link="logit")), color = "slateblue", fill = "lightskyblue2") + # Geom smooth can use the model to create curve and CI
  ylab ("Occurrence Probability") +
  xlab ("Grassland Core Area (ha)") +
  theme_bw()
# Create a graph with male/female (1/0) on y and wing chord on x 

grsp_ha_binomial_graph #Call object 
```
```{r abundance nonzero histogram, warning=FALSE, echo=FALSE, message=FALSE, fig.show = "hold", out.width = "85%", fig.align = "center", fig.cap= "Figure 4. Histogram showing counts of different grasshopper sparrow abundances for all nonzero values"}
ggplot(grsp_abundance_nonzero_data, aes(x=grsp)) + geom_histogram(binwidth=1, colour="black", fill="skyblue")+
  xlab("Grasshopper Sparrow Abundance") +
  ylab("Count")+
  theme_bw()

```


# Binomial Model


# Nonzero Poisson GLM
